From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-04 Pi+/paper/ufin.tex
Update section titles
Move some text in UFin

\section{The (Sub)Universe of Finite Types}
\label{sec:ufin}

@@

To define the weak groupoid we're after, we first introduce some basic concepts and notation from Homotopy Type Theory
that we use.~\footnote{We use the language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
that is, we use intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive
Types (HITs) for propositional truncation and set quotients. All arguments will hold in a Cubical Type Theory as well,
such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
For more details about HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}. By formalising our results in HoTT, we're also able to
use a proof assistant to computer-check our proofs, and our results are formalised in the Agda proof assistant using the
HoTT-Agda library. Using the formalisation of our semantics, we are able to extract a computable procedure for
performing NbE and synthesis of reversible circuits, as motivated in~\cref{sec:introduction}.

The weak groupoid structure of types is given by the identity type in HoTT, where points are given by terms of the type,
and higher morphisms are given by the iterated identity type. Functions between types are functors between groupoids,
and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to \UU$ comes
equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the structure of a
fibration (see the appendix for details).

@@

A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
function extensionality. An equivalent between types $A \eqv B$ is given by a pair of functions between them which
compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$. This is equivalent to the
identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

@@

>>>>

\section{The Groupoid of Finite Types}
\label{sec:ufin}

@@

We review some basic concepts and notation from Homotopy Type Theory that we use in the appendix.~\footnote{We use the
language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional
Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for propositional
truncation and set quotients. All arguments will hold in a Cubical Type Theory as well, such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
For more details about HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

To recap, types in HoTT have a weak groupoid structure given by the identity type, where points are given by terms of
the type, and higher morphisms are given by the iterated identity type. Functions between types are functors between
groupoids, and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to
\UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the
structure of a fibration (see the appendix for details).

@@

A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
function extensionality. An equivalence between types $A \eqv B$ is given by a pair of functions between them which
compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$, and this is equivalent to the
identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.

@@

\subsection{Univalent Fibrations}
